Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-03-28 paper/paper.tex
More things

\begin{abstract}
    The maximum common subgraph problem is to find the largest subgraph common to two given graphs.
    This problem can be solved either by constraint-based search, or by reduction to the maximum
    clique problem. We evaluate these two approaches using modern algorithms. We also propose a
    restricted branching approach for the connected version of the problem,
    show how to implement this in constraint-based and clique-inspired
    algorithms, and compare it to conventional propagation.
\end{abstract}

@@

Max common subgraph ?? cite some history and why we care

Connected \cite{DBLP:journals/tcs/Koch01,DBLP:conf/mco/VismaraV08}

CP models \cite{DBLP:conf/cp/NdiayeS11}

Clique models, discuss below.

@@

@@

\begin{figure}[p]
    \centering
    \input{gen-graph-unconnected-33}
    \caption{Clique versus CP, 33\% labelled. On the top, the cumulative
    number of instances solved by each approach, in under a certain time. On the bottom, the best
    clique approach compared to the best CP approach: the colour at a point indicates the number of
    instances with these runtimes.} \label{figure:unconnected-33}
\end{figure}

\begin{figure}[p]
    \centering
    \input{gen-graph-unconnected-plain}
    \caption{Clique versus CP, undirected, unlabelled. On the top, the cumulative
    number of instances solved by each approach, in under a certain time. On the bottom, the best
    clique approach compared to the best CP approach: the colour at a point indicates the number of
    instances with these runtimes.} \label{figure:unconnected-plain}
\end{figure}

@@

\section{Maximum Common Connected Subgraph}

@@

The key idea is that whenever we branch (except for the first choice we make), we select only from
variables which correspond to vertices which are adjacent to at least one other variable which has
been instantiated to a non-null value. If no such variables remain, the subgraph we are building
cannot grow any further, and all remaining variables must be instantiated to null.

?? We only need to work off whichever input gives us the variables. But the CP approach seems to
look at both?

@@

The $\mathit{branchable}$ set contains the set of association graph vertices which correspond to
vertices adjacent to an already-accepted vertex in the first input graph---in CP terms, it is the
set of assignments which could be made next which preserve connectedness. (Using only one of the two
input graphs is sufficient for correctness, and has the advantage that the $\FuncSty{connected}$
function may be implemented as a simple lookup into a precomputed array which maps each vertex in
the first input graph to a bitset.) At the top of search, this set is empty, and is not used (our
first vertex selection is special, and does not care about connectedness). At subsequent depths, we
may only accept vertices which are in this set, and if no such vertices remain then we return
immediately (\lineref{acceptbranchable}). When recursing, we extend $\mathit{branchable}$ with the
new vertices permitted by our acceptance of the branching $v$ (\lineref{addtobranchable}). Note that
we are assuming that inside the main loops, we encounter every vertex in $\mathit{remaining} \cap
\mathit{connected}$ before any vertex in $\mathit{remaining} \setminus \mathit{connected}$.

@@

\paragraph{Connected, Undirected, 33\% Labelled} \cref{figure:connected-undir33} Above two seconds,
    the clique approach dominates except for three instances, and there is only one instance which
    the CP model can solve but the clique model cannot (?? is this an EHP and does parallel get rid
    of it?). The clique model is unable to solve only ??  instances within one hour, ?? all of which
    have 90 or 100 vertices.  On the other hand, for trivial instances, the clique approach is often
    expensive due to the association graph construction.

@@

\begin{figure}[p]
    \centering
    \input{gen-graph-connected-undir33}
    \caption{Clique versus CP, connected, undirected, 33\% labelled, up to 100 vertices. On the top,
        the cumulative number of instances solved by each approach, in under a certain time. On the
        bottom, the best clique approach compared to the best CP approach: the colour at a point
        indicates the number of instances with these runtimes.} \label{figure:connected-undir33}
\end{figure}

\begin{figure}[p]
    \centering
    \input{gen-graph-connected-plain}
    \caption{Clique versus CP, connected, undirected, unlabelled, up to 35 vertices. On the top, the
    cumulative number of instances solved by each approach, in under a certain time. On the bottom,
    the best clique approach compared to the best CP approach: the colour at a point indicates the
    number of instances with these runtimes.}
\label{figure:connected-plain}
\end{figure}

>>>>

\begin{abstract}
    The maximum common subgraph problem is to find the largest subgraph common to two given graphs.
    This problem can be solved either by constraint-based search, or by reduction to the maximum
    clique problem. We evaluate these two approaches using modern algorithms. We also investigate a
    restricted branching approach for the connected version of the problem, show how to implement
    this in constraint-based and clique-inspired algorithms, and compare it to conventional
    propagation.
\end{abstract}

@@

\begin{figure}[b] % first page if possible
    \centering
    \begin{tikzpicture}[scale=0.33]%{{{
        \definecolor{uofgcobalt}{rgb}{0, 0.615686, 0.92549}
        \definecolor{uofgsunshine}{rgb}{1.0, 0.862745, 0.211765}
        \definecolor{uofgsandstone}{rgb}{0.321569, 0.278431, 0.231373}
        \begin{scope}[rotate=90]
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M1) at (90:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M2) at (150:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M3) at (30:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M4) at (210:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M5) at (330:1.5) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M6) at (270:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M7) at ($(210:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M8) at ($(330:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M9) at ($(270:1.5) + (M6)$) {\phantom{0}};

            \draw [very thick] (M1) -- (M2);
            \draw [very thick] (M2) -- (M4);
            \draw [very thick] (M3) -- (M5);
            \draw [color=uofgsandstone!40] (M4) -- (M6);
            \draw [color=uofgsandstone!40] (M5) -- (M6);
            \draw [very thick] (M3) -- (M1);
            \draw [color=uofgsandstone!40] (M6) -- (M7);
            \draw [color=uofgsandstone!40] (M6) -- (M8);
            \draw [color=uofgsandstone!40] (M6) -- (M9);
            \draw [very thick] (M7) -- (M9);
            \draw [very thick] (M8) -- (M9);
        \end{scope}

        \begin{scope}[xshift=7cm, rotate=90]
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M1) at (90:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M2) at (150:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M3) at (30:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M4) at (210:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M5) at (330:1.5) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M6) at (270:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M7) at ($(210:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M8) at ($(270:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M9) at ($(330:1.5) + (M8)$) {\phantom{0}};

            \draw [very thick] (M1) -- (M2);
            \draw [very thick] (M2) -- (M4);
            \draw [very thick] (M3) -- (M5);
            \draw [color=uofgsandstone!40] (M4) -- (M6);
            \draw [color=uofgsandstone!40] (M5) -- (M6);
            \draw [very thick] (M3) -- (M1);
            \draw [color=uofgsandstone!40] (M6) -- (M7);
            \draw [very thick] (M7) -- (M8);
            \draw [very thick] (M8) -- (M9);
        \end{scope}

        \begin{scope}[xshift=18cm, rotate=90]
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M1) at (90:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M2) at (150:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M3) at (30:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M4) at (210:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M5) at (330:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M6) at (270:1.5) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M7) at ($(210:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M8) at ($(330:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M9) at ($(270:1.5) + (M6)$) {\phantom{0}};

            \draw [very thick] (M1) -- (M2);
            \draw [very thick] (M2) -- (M4);
            \draw [very thick] (M3) -- (M5);
            \draw [very thick] (M4) -- (M6);
            \draw [very thick] (M5) -- (M6);
            \draw [very thick] (M3) -- (M1);
            \draw [color=uofgsandstone!40] (M6) -- (M7);
            \draw [very thick] (M6) -- (M8);
            \draw [color=uofgsandstone!40] (M6) -- (M9);
            \draw [color=uofgsandstone!40] (M7) -- (M9);
            \draw [color=uofgsandstone!40] (M8) -- (M9);
        \end{scope}

        \begin{scope}[xshift=25cm, rotate=90]
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M1) at (90:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M2) at (150:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M3) at (30:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M4) at (210:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M5) at (330:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M6) at (270:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (M7) at ($(210:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M8) at ($(270:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M9) at ($(330:1.5) + (M8)$) {\phantom{0}};

            \draw [very thick] (M1) -- (M2);
            \draw [very thick] (M2) -- (M4);
            \draw [very thick] (M3) -- (M5);
            \draw [very thick] (M4) -- (M6);
            \draw [very thick] (M5) -- (M6);
            \draw [very thick] (M3) -- (M1);
            \draw [very thick] (M6) -- (M7);
            \draw [draw=uofgsandstone!40] (M7) -- (M8);
            \draw [draw=uofgsandstone!40] (M8) -- (M9);
        \end{scope}
\end{tikzpicture}
\caption{A maximum common induced subgraph of the first two graphs has eight vertices. There are
four symmetric solutions, all involving the shaded vertices. However, if we require the common
subgraph to be connected, only seven vertices may be selected---one way to do this is shown in the
third and fourth graphs.}\label{figure:mcsexample}
\end{figure}

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, computer vision ?? what should we cite?, ??
anything else?, both directly and as a way of measuring the similarity or difference between two
graphs \cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}. We illustrate
this in \cref{figure:mcsexample}, and define the problem formally below.

@@

?? Get edge vs vertex in early.

\subsection{Existing Approaches}

One approach to maximum common subgraph problems is via constraint programming
\cite{DBLP:conf/mco/VismaraV08,DBLP:conf/cp/NdiayeS11}. The models are based upon constructing an
injective mapping from the vertices of one graph (typically the smaller) to the other, in a similar
manner to subgraph isomorphism algorithms
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}. However, to
allow for a partial mapping, a special $\bot$ value is added to each domain. The objective is then
to minimise the number of variables given the $\bot$ value---for best results, a special propagator
for the injectivity condition is used, which simultaneously enforces that all variables taking
non-$\bot$ are different, and restricts the number of $\bot$ values which may be used based upon the
objective \cite{DBLP:conf/cp/PetitRB01}.  Unfortunately, the addition of $\bot$ also invalidates
many of the invariants used for subgraph isomorphism---we may no longer reason using the degree of
vertices, or the distances between them, for example, and so maximum common subgraph problems can be
much more challenging in practice.

Another widely studied approach is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover) in an association graph (which we describe
below). Conventional wisdom is that this approach is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph approach using a modern
algorithm.

(Other approaches have been tried, including mixed integer programming
\cite{DBLP:journals/anor/PivaS12}; SAT encodings seem to struggle even for subgraph isomorphism??
cite IJCAI if it's not rejected?.)

An interesting variant of the maximum common subgraph problem requires the selected subgraph to be
connected
\cite{DBLP:journals/tcs/Koch01,DBLP:journals/jcamd/RaymondW02a,DBLP:conf/mco/VismaraV08,Ehrlich:2011}.
Adding the connectedness requirement makes certain special cases solvable in polynomial time,
including outerplanar graphs of bounded degree \cite{DBLP:journals/algorithms/AkutsuT13} and trees
\cite{DBLP:journals/corr/DroschinskyKM16}, but the general case remains \NP-hard. The second half of
this paper looks at this variant on arbitrary graphs. In principle, for a constraint programming
approach, we may simply add a ``forms a connected graph'' global constraint to the model
\cite{Brown:2005,DBLP:conf/cp/DoomsDD05,DBLP:conf/cp/QuesadaRD05}. Alternatively, we could use a
special branching rule to maintain connectedness during search ?? cite. We see that using both
approaches together gives best results ?? does it?.  We also explain how to adapt this branching
rule to run inside a modern clique-like algorithm, to solve the problem via an association graph.

?? Summary, association graph is better when edges have labels.

@@

\begin{figure}[p]
    \centering
    \input{gen-graph-unconnected-33}
    \caption{Clique versus constraint programming, 33\% labelled. On the top, the cumulative
    number of instances solved by each approach, in under a certain time. On the bottom, the best
    clique approach compared to the best constraint programming approach: the colour at a point indicates the number of
    instances with these runtimes.} \label{figure:unconnected-33}
\end{figure}

\begin{figure}[p]
    \centering
    \input{gen-graph-unconnected-plain}
    \caption{Clique versus constraint programming, undirected, unlabelled. On the top, the cumulative
    number of instances solved by each approach, in under a certain time. On the bottom, the best
    clique approach compared to the best constraint programming approach: the colour at a point indicates the number of
    instances with these runtimes.} \label{figure:unconnected-plain}
\end{figure}

\subsection{Edge Labels}

@@

\section{Find Maximum Common Connected Subgraphs}

@@

The key idea is that whenever we branch (except for the first choice we make), we select only from
variables which correspond to vertices which are adjacent to at least one other variable which has
been instantiated to a value other than $\bot$. If no such variables remain, the subgraph we are building
cannot grow any further, and all remaining variables must be instantiated to $\bot$.

?? We only need to work off whichever input gives us the variables. But the constraint programming approach seems to
look at both?

@@

The $\mathit{branchable}$ set contains the set of association graph vertices which correspond to
vertices adjacent to an already-accepted vertex in the first input graph---in constraint programming terms, it is the
set of assignments which could be made next which preserve connectedness. (Using only one of the two
input graphs is sufficient for correctness, and has the advantage that the $\FuncSty{connected}$
function may be implemented as a simple lookup into a precomputed array which maps each vertex in
the first input graph to a bitset.) At the top of search, this set is empty, and is not used (our
first vertex selection is special, and does not care about connectedness). At subsequent depths, we
may only accept vertices which are in this set, and if no such vertices remain then we return
immediately (\lineref{acceptbranchable}). When recursing, we extend $\mathit{branchable}$ with the
new vertices permitted by our acceptance of the branching $v$ (\lineref{addtobranchable}). Note that
we are assuming that inside the main loops, we encounter every vertex in $\mathit{remaining} \cap
\mathit{connected}$ before any vertex in $\mathit{remaining} \setminus \mathit{connected}$.

@@

\paragraph{Connected, Undirected, 33\% Labelled} \cref{figure:connected-undir33} Above two seconds,
    the clique approach dominates except for three instances, and there is only one instance which
    the constraint programming model can solve but the clique model cannot (?? is this an EHP and does parallel get rid
    of it?). The clique model is unable to solve only ??  instances within one hour, ?? all of which
    have 90 or 100 vertices.  On the other hand, for trivial instances, the clique approach is often
    expensive due to the association graph construction.

@@

\begin{figure}[p]
    \centering
    \input{gen-graph-connected-undir33}
    \caption{Clique versus constraint programming, connected, undirected, 33\% labelled, up to 100 vertices. On the top,
        the cumulative number of instances solved by each approach, in under a certain time. On the
        bottom, the best clique approach compared to the best constraint programming approach: the colour at a point
        indicates the number of instances with these runtimes.} \label{figure:connected-undir33}
\end{figure}

\begin{figure}[p]
    \centering
    \input{gen-graph-connected-plain}
    \caption{Clique versus constraint programming, connected, undirected, unlabelled, up to 35 vertices. On the top, the
    cumulative number of instances solved by each approach, in under a certain time. On the bottom,
    the best clique approach compared to the best constraint programming approach: the colour at a point indicates the
    number of instances with these runtimes.}
\label{figure:connected-plain}
\end{figure}
